# IMDb Movie Reviews

## Learning objectives

* Identify and create appropriate summary statistics and plots
  given a data set or research question for quantitative data

* Interpret the following summary statistics in context:
  median, lower quartile, upper quartile,
  standard deviation, inter-quartile range

* Given a plot or set of plots, describe and compare the distribution(s)
  of a single quantitative variable
  (center, variability, shape, outliers)

## Terminology review

In this week's activity, we will review summary measures and plots for quantitative variables.  Some terms covered in this activity are:

* Two measures of center: mean, median

* Two measures of spread (variability): standard deviation, inter-quartile range (IQR)

* Types of graphs: box plots, dot plots, histograms

To review these concepts, see Section 2.3 in the textbook.

## Movies released in 2016

A data set was collected on movies released in 2016.  Here is a list of some of the variables collected on these movies.

| **Variable** 	| **Description** |
|----	|-------------	|
| `budget_mil` | Amount of money (in US $ millions) budgeted for the production of the movie |
| `revenue_mil` | Amount of money (in US $ millions) the movie made after release|
| `duration` | Length of the movie (in minutes)|
| `content_rating` | Rating of the movie (`G`, `PG`, `PG-13`, `R`, `Not Rated`)|
| `imdb_score` | IMDb user rating score from 1 to 10 |
| `genres` | Categories the movie falls into (e.g., Action, Drama, etc.) |
| `movie_facebook_likes` | Number of likes a movie receives on Facebook |

\newpage

### Vocabulary review. Complete Q1 - 3 before class. {-}

1. What are the observational units in this data set?

\vspace{0.1in}

2. Which of the above listed variables are categorical?

\vspace{.5in}

3. Which of the above listed variables are quantitative?

\vspace{.5in}

### Summarizing a single quantitative variable {-}

The `favstats` function from the `mosaic` package gives the summary statistics for a quantitative variable. Here we have the summary statistics for the variable `imdb_score`.

```{r, echo=TRUE, collapse = FALSE}
movies <- read.csv("data/Movies2016.csv") # Read in data set
movies %>% #Data set piped into...
  summarise(favstats(imdb_score)) #Apply favstats function to imdb_score
```

4. Give the values for the two measures of center.

\vspace{0.5in}

5. Calculate the Interquartile range (IQR = Q3 - Q1).

\vspace{0.5in}

6. Report the value of the standard deviation and interpret this value in context of the problem.
\vspace{1in}

### Displaying a single quantitative variable {-}

7. What are the three types of plots used to plot a single quantitative variable?

\newpage

To create a histogram of the IMDb scores, enter the variable name, `imdb_score` in the provided `RScript` file for xx at line 12, highlight and run lines 1 - 16.  Visually, this shows us the range of IMDb scores for Movies released in 2016.

Notice that the **bin width** is 0.5.  For example the first bin consists of the number of movies in the data set with an IMDb score of 3.25 to 3.75.  It is important to note that a movie with a IMDb score on the boundary of a bin will fall into the bin above it; for example, 4.76 would be counted in the bin 4.75--5.25.  

```{r, out.width="60%", echo=TRUE, eval=FALSE}
movies %>% #Data set piped into...
ggplot(aes(x = xx)) +   #Name variable to plot
  geom_histogram(binwidth = 0.5) +  #Create histogram with specified binwidth
  labs(title = "Histogram of IMDb Score of Movies in 2016", #title for plot
       x = "IMDb Score", #Label for x axis
       y = "Frequency") #Label for y axis
```


8. Sketch the histogram created here.

\vspace{1in}

9. Which range of IMDb scores have the highest frequency?

\vspace{0.4in}

10. What is the shape of the distribution of IMDb scores?

\vspace{0.4in}

11. Which five summary statistics are used in creating a box plot? *Hint*: Together they are called the **five-number summary** of the variable.

\vspace{0.4in}

12. Using the code below we see that the three smallest IMDb scores in the data set are 3.4, 3.5, and 3.7 and the three largest IMDb scores are 8.0, 8.1, and 8.2:  

```{r, indent = "    ", echo = TRUE, collapse = FALSE}
movies %>% # Data set pipes into...
  select(imdb_score) %>% # Select imdb_score variable
  slice_min(imdb_score, n = 3)  # Show 3 smallest values
```

```{r, indent = "    ", echo = TRUE, collapse = FALSE}
movies %>% # Data set pipes into...
  select(imdb_score) %>% # Select imdb_score variable
  slice_max(imdb_score, n = 3)  # Show 3 largest values
```

    Using the summary statistics above, and the smallest and largest values of the variable to check for outliers, sketch a box plot of IMDb Score.  Be sure to label the axes.

\vspace{1.5in}

### Displaying a single categorical and single quantitative variable {-}

The box plot of movie budgets (in millions) by content rating is plotted using the code below.  Enter the variable `budget_mil` for yy and the variable `content_rating` for xx at line 31, highlight and run code lines 29 - 35. This plot helps to compare the budget for different levels of content rating.

```{r, out.width="60%", echo=TRUE, eval=FALSE}
movies %>%  #Data set piped into...
  filter(content_rating != "Not Rated") %>% # Remove Not Rated movies
  ggplot(aes(y = yy, x = xx))+  #Identify variables
  geom_boxplot()+  #Tell it to make a box plot
  labs(title = "Side by side box plot of budget by content rating",  #Title
       x = "Content Rating",    #x-axis label
       y = "Budget (in Millions)")  #y-axis label
```

13. Sketch the boxplots created using the `R` code.

\vspace{1.5in}

\newpage

14. Answer the following questions about the box plots created.

   a. Which content rating has the highest center?
\vspace{0.2in}

   b. Which content rating has the largest spread?
\vspace{0.2in}

   c. Which content rating has the most skewed distribution?
\vspace{0.2in}

   d. Fifty percent of movies in 2016 with a PG-13 content rating fall below what value?  What is the name of this value?
\vspace{0.4in}

   e.  What is the value for the third quartile (Q3) for the PG-13 rating?  Interpret this value in context.
\vspace{.8in}

## Out of Class Activity
For a little more practice using `Rstudio` to create graphs of quantitative variables we will look at some other variables in the `Movies` data set.  Download and open the provided `RScript` file, highlight and run the first 8 lines of code.

To use the favstats function in the mosaic package with two variables, we will enter the variables as a formula, response~explanatory.

```{r, echo=TRUE, collapse=FALSE}
movies %>% #Data set piped into...
  summarise(favstats(imdb_score~content_rating)) #Apply favstats function to imdb_score
```

Using the provided `RScript` file, we will create side-by-side histograms of IMDb by movie content rating. Enter the variable name, `imdb_score` for yy and `content_rating` for xx at line 44, highlight and run lines 39 - 48.

```{r, out.width="60%", echo=TRUE, eval=FALSE}
movies %>%  #Data set piped into...
  filter(content_rating != "Not Rated") %>% # Remove Not Rated movies
  ggplot(aes(y = yy, x = xx))+  #Identify variables
  geom_boxplot()+  #Tell it to make a box plot
  labs(title = "Side by side box plot of budget by content rating",  #Title
       x = "Content Rating",    #x-axis label
       y = "Budget (in Millions)")  #y-axis label
```

1.  Using the provided favstats output and the side-by-side boxplots, interpret the value of quartile 1 for the R content rating.

\vspace{1in}

2. Which content rating has the highest center?

\vspace{0.2in}

3. Which variable is the explanatory variable? Response variable?

\vspace{0.5in}

## Additional notes

Use this space to summarize your thoughts and take additional notes on this week's activity and material covered.