## Week 4 Lab:  Penguins

\setstretch{1}

### Learning outcomes

* Identify and create appropriate summary statistics and plots
  given a data set with two quantitative variables.
  
* Use scatterplots to assess the relationship between two quantitative variables.

* Find the estimated line of regression using summary statistics and R linear model (`lm()`) output.

* Interpret the slope coefficient in context of the problem.

* Calculate and interpret $R^2$, the coefficient of determination, in context of the problem.

* Find the correlation coefficient from R output or from $R^2$ and the sign of the slope.

### Penguins

The Palmer Station Long Term Ecological Research Program sampled three penguin species on islands in the Palmer Archipelago in Antarctica. Researchers took various body measurements on the penguins, including flipper length and body mass. The researchers were interested in the relationship between flipper length and body mass and wondered if flipper length could be used to accurately predict the body mass of these three penguin species. 

Upload and import the `Antarctica_Penguins` csv file and the provided R script file for week 4 lab. Enter the name of the data set (see the environment tab) for `datasetname` in the R script file in line 4.

First we will create a scatterplot of the flipper length and body mass.  Notice that we are using flipper length to predict body mass.  This makes flipper length the explanatory variable. **Make sure to give your plot a descriptive title.** Highlight and run lines 1--13 in the R script file.  **Upload a copy of your scatterplot to Gradescope.**

```{r, out.width="60%", echo=TRUE, eval=FALSE}
penguins <- datasetname #Creates the object penguins
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g))+  # Specify variables
  geom_point() +  # Add scatterplot of points
  labs(x = "flipper length (mm)",  # Label x-axis
       y = "body mass (g)",  # Label y-axis
       title = "Title") + # Be sure to title your plots
  geom_smooth(method = "lm", se = FALSE)  # Add regression line
```

1. Assess the four features of the scatterplot that describe this relationship.
\vspace{1mm}

* Form (linear, non-linear)

\vspace{.1in}

* Direction (positive, negative)

\vspace{.1in}

* Strength

\vspace{.1in}

* Unusual observations or outliers

\vspace{.1in}

Highlight and run lines 16--20 to get the correlation matrix in the R script file.

```{r, echo=TRUE, collapse = FALSE, eval=FALSE}
penguins %>%  # Data set pipes into
  select(c("bill_length_mm", "bill_depth_mm", 
           "flipper_length_mm", "body_mass_g")) %>%
  cor(use="pairwise.complete.obs") %>%
  round(3)
```

2.  Using the R output, which two variables have the *strongest* correlation? What is the value of this correlation?

\vspace{0.5in}

3. Using the value of correlation found in question 2, calculate the value of the coefficient of determination.

\vspace{0.5in}

4. **Interpret the coefficient of determination in context of the problem.**

\vspace{1in}

Enter the variable `body_mass_g` for `response` and the variable name `flipper_length_mm` for `explanatory` in line 23 in the R script file.  Highlight and run lines 23--24 to get the linear model output.

```{r, echo=TRUE, collapse = FALSE, eval=FALSE}
# Fit linear model: y ~ x
penguinsLM <- lm(response~explanatory, data=penguins)
summary(penguinsLM)$coefficients # Display coefficient summary
```

5.  Write out the least squares regression line using the summary statistics from the R output in context of the problem.

\vspace{.5in}

6. **Interpret the value of slope in context of the problem.**

\vspace{.8in}

7. **Using the least squares regression line from question 5, predict the body mass for a penguin with a flipper length of 181 mm.**

\vspace{.6in}

8. One penguin had a flipper length of 181 mm and a body mass of 3750 g. Find the residual for this penguin.

\vspace{.8in}

9.  Did the line of regression overestimate or underestimate the body mass for this penguin?

\vspace{0.5in}

Highlight and run lines 27--34 to get the multivariate plot.

```{r, out.width="60%", echo=TRUE, eval=FALSE}
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color=species))+  # Specify variables
  geom_point(aes(shape = species), size = 3) +  # Add scatterplot of points
  labs(x = "flipper length (mm)",  # Label x-axis
       y = "body mass (g)",  # Label y-axis
       color = "species",
       title = "TITLE") + # Be sure to tile your plots
  geom_smooth(method = "lm", se = FALSE)  # Add regression line
```

10.  What three variables are plotted on this plot?

\vspace{0.3in}


11. **Does adding the variable species affect the relationship between body mass and flipper length? Explain your answer.**

\newpage

