# Movie Profits

## Learning objectives

* Identify and create appropriate summary statistics and plots
  given a data set with two quantitative variables
  
* Use scatterplots to assess the relationship between two quantitative variables

* Find the correlation coefficient 

* Find the estimated line of regression using summary statistics and `R` linear model (`lm`) output

* Interpret the slope coefficient in context of the problem

* Interpret the coefficient of determination in context of the problem

## Terminology review

In today's activity, we will review summary measures and plots for two quantitative variables.  Some terms covered in this activity are:

* Scatterplot

* Correlation

* Slope

* Least-squares line of regression

* Coefficient of determination ($r$-squared)

To review these concepts, see Chapter 3 in the textbook.  

## Movies released in 2016

We will revisit the data set used last week collected on Movies released in 2016. Here is a reminder of the variables collected on these movies.

| **Variable** 	| **Description** |
|----	|-------------	|
| `budget_mil` | Amount of money (in US $ millions) budgeted for the production of the movie |
| `revenue_mil` | Amount of money (in US $ millions) the movie made after release|
| `duration` | Length of the movie (in minutes)|
| `content_rating` | Rating of the movie (`G`, `PG`, `PG-13`, `R`, `Not Rated`)|
| `imdb_score` | IMDb user rating score from 1 to 10 |
| `genres` | Categories the movie falls into (e.g., Action, Drama, etc.) |
| `movie_facebook_likes` | Number of likes a movie receives on Facebook |

\newpage

### Vocabulary review {-}

1.  What type of plot is used to display two quantitative variables?

\vspace{0.2in}

2.  What summary statistics are used to describe the relationship between two quantitative variables?

\vspace{0.3in}

We will look at the relationship between 'Budget' and 'Revenue' for movies released in 2016. This shows a scatterplot of 'Budget' as a predictor of 'Revenue' (Note: both variables are measures in "millions of dollars").  

```{r echo=t, fig.height=5,fig.width=12, message = FALSE}
movies <- read.csv("data/Movies2016.csv") #Reads in data set
movies %>% #Data set pipes into...
ggplot(aes(x = budget_mil, y = revenue_mil))+  #Specify variables
  geom_point() +  #Add scatterplot of points
  labs(x = "Budget in Millions ($)",  #Label x-axis
       y = "Revenue in Millions ($)",  #Label y-axis
       title = "Revenue vs. Budget") + #Be sure to tile your plots
  geom_smooth(method = "lm", se = FALSE)  #Add regression line
```
3. Assess the four features of the scatterplot that describe this relationship. Describe each feature using a complete sentence!

* Form (linear, non-linear)

\vspace{.4in}

* Direction (positive, negative)

\vspace{.4in}

* Strength

\vspace{.4in}

* Unusual observations or outliers

\vspace{.4in}

\newpage

4. Does there appear to be an association between 'Budget' and 'Revenue'? Explain.

\vspace{1in}

### Correlation  {-}

Correlation measures the strength and the direction between two quantitative variables.  The closer the value of correlation to + or - 1 the stronger the linear relationship.  Values close to zero indicate a very weak linear relationship between the two variables.  The following output shows a correlation matrix between several pairs of quantitative variables.  

```{r, echo=TRUE, collapse = FALSE}
# Take subset of variables
movies %>%  #Data set pipes into...
  select(c("budget_mil", "revenue_mil",  # Take subset of variables
           "duration", "imdb_score", 
           "movie_facebook_likes")) %>%
  cor(use="pairwise.complete.obs") %>% # Calculate correlation matrix
  round(3)  # Round to 3 decimals
```

5.  Using the output above, which two variables have the strongest correlation?

\vspace{0.3in}

6.  What is the value of correlation between 'Budget' and 'Revenue'?

\vspace{0.3in}

7.  Based on the value of correlation what would the sign of the slope be? Positive or negative?  Explain.

\vspace{0.5in}

8.  Does your answer to question 7 match the direction you choose in question 3?

\vspace{0.2in}

9.  Explain why the correlation values on the diagonal are equal to 1.

\vspace{1in}

\newpage

### Slope {-}

The linear model function in `R` gives us the summary for the least squares regression line.  The estimate for `(Intercept)` is the $y$-intercept for the line of least squares and the estimate for `budget_mil` is the value of $b_1$, the slope.

```{r, echo=TRUE, collapse = FALSE}
# Fit linear model: y ~ x
revenueLM <- lm(revenue_mil ~ budget_mil, data=movies)
summary(revenueLM)$coefficients # Display coefficient summary
```

You may remember from middle and high school that slope $=\frac{\mbox{rise}}{\mbox{run}}$.  

Using $b_1$ to represent slope, we can write that as the fraction $\frac{b_1}{1}$. 

Therefore, the slope predicts the how much the line will *rise* for each *run* of +1. In other words, as the $x$ variable increases by 1 unit, the $y$ variable is expected to change (increase/decrease) by the value of slope.


10.  Write out the least squares line using the summary statistics provided in proper statistical notation.

\vspace{.6in}

11. Interpret the value of slope in context of the problem.

\vspace{1in}

12. Using the least squares line from question 10, predict the revenue for a movie with a budget of 165 million.

\vspace{.6in}

### Residuals {-}

The model we are using assumes the relationship between the two variables follows a straight line. The residuals are the errors, or the part that hasn't been modeled by the line.

\begin{center}
Data = Model + Residual

Residual = Data - Model

$e_i=y_i-\hat{y}_i$
\end{center}

13.  The movie, *Independence Day: Resurgence*, had a budget of 165 million and revenue of 102.315 million.  Find the residual for this movie.

\vspace{.8in}

14.  Did the line of regression overestimate or underestimate the revenue for this movie? 

\vspace{.2in}

\newpage

### Coefficient of determination (squared correlation) {-}

The coefficient of determination, $r^2$, can also be used to describe the strength of the linear relationship between two quantitative variables. $r^2$ measures the proportion of variation in the response that is explained by the least squares line with the explanatory variable.  


15.  Use the correlation, $r$, to calculate the coefficient of determination between 'Budget' and 'Revenue', $r^2$.

\vspace{.4in}

16.  Interpret the coefficient of determination in context of the problem.

\vspace{.6in}

### Multivariate plots {-}
What if we wanted to see if the relationship between 'Budget' and 'Revenue' differs if we add another variable into the picture?  The following plot visualized three variables, creating a **multivariate** plot. 

```{r echo=TRUE, fig.height=5,fig.width=12, message = FALSE}
movies %>% #Data set pipes into...
ggplot(aes(x = budget_mil, y = revenue_mil, color = content_rating)) +  #Specify variables
  geom_point(aes(shape = content_rating), size = 3) +  #Add scatterplot of points
  labs(x = "Budget in Millions ($)",  #Label x-axis
       y = "Revenue in Millions ($)",  #Label y-axis
       color = "Content Rating",  #Label legend
       title = "Revenue vs. Budget") + #Be sure to tile your plots
  geom_smooth(method = "lm", se = FALSE, lwd = 2) + #Add regression lines
  scale_color_grey() #Make black and white
```

17.  Identify the three varables plotted in this graph.

\vspace{0.5in}

18. Does the relationship between 'Budget' and 'Revenue' differ among the different content ratings?  Explain.

\vspace{1in}

\newpage

## Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity.


