## Activity 7B:  Handedness of Male Boxers --- Theory-based Methods

\setstretch{1}

### Learning objectives

* Describe and perform a theory-based hypothesis test for a single proportion.

* Check the appropriate conditions to use a theory-based methods.

* Calculate and interpret the standardized sample proportion.

* Interpret and evaluate a p-value for a theory-based hypothesis test for a single proportion.

* Use the normal distribution to find the p-value.

* Calculate a theory-based confidence interval for a single proportion.

* Interpret a confidence interval for a single proportion.

* Use the normal distribution to find the multiplier needed for a confidence interval

### Terminology review

In this activity, we will introduce theory-based hypothesis tests and confidence intervals for a single proportion. Some terms covered in this activity are:

* Parameter of interest

* Standardized Statistic

* Normal distribution

* p-value

* Multiplier

* Normal distribution

To review these concepts, see Chapters 11 & 14 in your textbook.


Activities 6A, 6B, and 7A covered simulation-based methods for hypothesis tests involving a single categorical variable. This activity covers theory-based methods for testing a single categorical variable.  

### Handedness of male boxers

Left-handedness is a trait that is found in about 10\% of the general population. Past studies have shown that left-handed men are over-represented among professional boxers [@richardson2019]. The fighting claim states that left-handed men have an advantage in competition.  In this random sample of 500 male professional boxers, we want to see if there is an over-prevalence of left-handed fighters.  In the sample of 500 male boxers, 81 were left-handed.

```{r, echo=TRUE, collapse=FALSE, message=FALSE, warning=FALSE}
 # Read in data set
boxers <- read.csv("https://math.montana.edu/courses/s216/data/Male_boxers_sample.csv")
boxers %>% count(Stance)  # Count number in each Stance category
```

### Review of summary statistics {-}

1.  Write out the parameter of interest for this study.  

\vspace{0.6in}

2.  Write out the null hypothesis in words.
\vspace{0.6in}

3. Write out the alternative hypothesis in notation.
\vspace{0.3in}

4. Give the value of the summary statistic (sample proportion) for this study.  Use proper notation.

\vspace{0.3in}

### Theory-based methods {-}

The sampling distribution of a single proportion --- how that proportion varies from sample to sample --- can be mathematically modeled using the normal distribution if certain conditions are met.

Conditions for the sampling distribution of $\hat{p}$ to follow an approximate normal distribution:

* **Independence**: The sample’s observations are independent, e.g., are from a simple random sample. (*Remember*: This also must be true to use simulation methods!)

* **Success-failure condition**: We *expect* to see at least 10 successes and 10 failures in the sample, $n\hat{p}≥10$  and $n(1-\hat{p})≥10$.

5. Verify that the independence condition is satisfied.

\vspace{0.8in}

6. Is the success-failure condition met to model the data with the normal distribution?  Show your work to support your answer.

\vspace{1in}
\newpage

To calculate the standardized statistic we use the general formula 

$$
Z = \frac{\text{point estimate} - \text{null value}}{SE_0(\text{point estimate})}.
$$
For a single categorical variable the standardized sample proportion is calculated using

$$
Z = \frac{\hat{p} - \pi_0}{SE_0(\hat{p})},
$$
where the standard error is calculated using the null value:

$$SE_0(\hat{p})=\sqrt{\frac{\pi_0(1-\pi_0)}{n}}$$.

The standard error of the sample proportion measures the variability of possible sample proportions from the actual proportion.  In other words, how far each possible sample proportion is from the actual proportion on average.  For this study, the null standard error of the sample proportion is calculated using the null value, 0.1.

$$SE_0(\hat{p})=\sqrt{\frac{0.1(1-0.1)}{500}} = 0.013$$.

7.  Interpret the null standard error of the sample proportion in context of the problem.

\vspace{0.6in}

8.  Using the null standard error of the sample proportion, calculate the standardized sample proportion.  

\vspace{0.6in}

\newpage

The standardized statistic is used as a ruler to measure how far the sample statistic is from the null value.  Essentially, we are converting the sample proportion into a measure of standard errors to compare to the standard normal distribution.  

9.  Using the 68-95-99.7 rule in Section 11.2.5 to guide you, fill in the percentages on the standard normal distribution displayed in Figure \@ref(fig:simpleNormalcurve), and also mark the value of the standardized statistic calculated in question 8.

```{r simpleNormalcurve, fig.cap="A standard normal curve.", warning=FALSE, out.width="50%"}
openintro::normTail(m = 0, s = 1)
```

The standardized statistic measures the *number of standard errors the sample statistic is from the null value*.

10.  Interpret the standardized sample proportion from question 8 in context of the problem.

\vspace{.8in}

We will use the `pnorm()` function in R to find the p-value. Use the provided R script file and enter the value of the standardized statistic calculated in question 8 at `xx` in line 7; highlight and run lines 7--9.  Notice that in line 9 it says `lower.tail = FALSE`.  R will calculate the p-value *greater* than the value of the standardized statistic.  

Notes:

* Use `lower.tail = TRUE` when doing a left-sided test.
* Use `lower.tail = FALSE` when doing a right-sided test.
* To find a two-sided p-value, use a left-sided test for negative Z or a right-sided test for positive Z, then multiply the value found by 2 to get the p-value.

```{r, echo=TRUE, eval=FALSE}
pnorm(xx, # Enter value of standardized statistic
      m=0, s=1, # Using the standard normal mean = 0, sd = 1
      lower.tail=FALSE) # Gives a p-value greater than the standardized statistic
```
  
  
11.  Report the p-value obtained from the R output.
\vspace{0.3in}

\newpage

12.  Write a conclusion based on the value of the p-value.

\vspace{0.5in}


#### Theory-based confidence interval {-}

To calculate a theory-based 95\% confidence interval for $\pi$, we will first find the **standard error** of $\hat{p}$ by plugging in the value of $\hat{p}$ for $\pi$ in $SD(\hat{p})$:

$$SE(\hat{p}) = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}.$$
Note that we do not include a "0" subscript, since we are not assuming a null hypothesis. 

13.  Calculate the standard error of the sample proportion to find a 95\% confidence interval.

\vspace{0.5in}

To find the confidence interval, we will add and subtract the **margin of error** to the point estimate:

$$\text{point estimate}\pm\text{margin of error}$$
$$\hat{p}\pm z^* \times SE(\hat{p})$$
$$ME = z^* \times SE(\hat{p})$$

The $z^*$ multiplier is the percentile of a standard normal distribution that corresponds to our confidence level. If our confidence level is 95\%, we find the Z values that encompass the middle 95\% of the standard normal distribution.  If 95\% of the standard normal distribution should be in the middle, that leaves 5\% in the tails, or 2.5\% in each tail.  
 

14. Fill in the normal distribution shown in figure 7.2 to show how R found the $z^*$ multiplier.

```{r simpleNormaldist, fig.cap="A standard normal curve.", warning=FALSE, out.width="50%"}
X <- seq(-5,5,0.01)
Y <- dnorm(X)
plot(X, Y, type='l', axes=F, xlim=c(-4,4), lwd=2, ylab="", xlab="Z")
#axis(1, at=-3:3)
abline(h=-0.002)
```

\newpage 

The `qnorm()` function in R will tell us the $z^*$ value for the desired percentile (in this case, 95\% + 2.5\% = 97.5\% percentile). Enter the value of 0.975 for xx in the provided R script file. This will give the value of the multiplier for a 95\% confidence interval.

```{r, echo=TRUE, eval=FALSE}
qnorm(xx) # Multiplier for 95% confidence interval
```

15. Report the value of the multiplier needed to calculate the 95\% confidence interval for the true proportion of male boxers that are left-handed?
\vspace{0.3in}

16.  Calculate the margin of error for the 95\% confidence interval.
\vspace{0.8in}

17.  Calculate the 95\% confidence interval for the parameter of interest.
\vspace{0.5in}

19.  Interpret the 95\% confidence interval in the context of the problem.
\vspace{0.8in}

20. Is the null value, 0.1, contained in the 95\% confidence interval?  Explain, based on the p-value from the earlier in activity, why you expected this to be true.
\vspace{0.5in}


\newpage

### Take-home messages

1.  In theory-based methods, we add and subtract a margin of error to the sample statistic.  The margin of error is calculated using a multiplier that corresponds to the level of confidence times the variability (standard error) of the statistic.

2. The confidence interval calculated using theory-based methods should be similar to the confidence interval found using simulation methods provided the success-failure condition is met.

<!-- 3.  A smaller sample size will increase the margin of error which results in a wider confidence interval.  -->

3. If repeat samples of the same size are selected from the population, approximately 95\% of samples will create a 95\% confidence interval that contains the parameter of interest.

### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity and material covered.


\newpage