## Week 7 Lab:  Comic Book Characters

\setstretch{1}

### Learning objectives

* Identify the two possible explanations (one assuming the null hypothesis, and one assuming the alternative hypothesis) for a relationship seen in sample data.

* Given a research question involving a single categorical variable, construct the null and alternative hypotheses in words and using appropriate statistical symbols.
  
* Describe and perform a simulation-based hypothesis test for a single proportion.

* Interpret and evaluate a p-value for a simulation-based hypothesis test for a single proportion.

* Use bootstrapping to find a confidence interval for a single proportion.

* Interpret a confidence interval for a single proportion.

* Interpret what confidence means in a confidence interval


### Comic Book Characters

Among all comic book characters produced by the comic book publisher Marvel, 25\% are female. The comic book publisher DC Comics has taken note of this disparity in the number of female versus male characters, and is trying to diversify their offerings. In a random sample of recently published DC Comics books, 52 characters were included in the comics, of which 15 are female.  **Remember that the bolded questions will be turned in on Gradescope for your group.**

1. What type of plot would be appropriate to display this data?

\vspace{0.3in}

2. Calculate the summary statistic.  Use appropriate notation.

\vspace{0.3in}

**3. Write the parameter of interest in context of the problem.**

\vspace{0.8in}

4.  Write the null hypothesis in notation.

\vspace{0.3in}

5. Write the alternative hypothesis in words. 

\vspace{0.8in}

6. Is the independence condition met?  Explain.

\vspace{0.6in}

7. What values should be entered for each of the following into the one proportion test to create 1000 simulations?

\vspace{1mm}

* Probability of success:

\vspace{.2in}
* Sample size:
    
\vspace{.2in}
* Number of repetitions:
    
\vspace{.2in}
* As extreme as:
    
\vspace{.2in}
* Direction (`"greater"`, `"less"`, or `"two-sided"`):

\vspace{.2in}

We will use the `one_proportion_test()` function in `R` (in the `catstats` package) to simulate the null distribution of sample proportions and compute a p-value. Using the provided `R` script file, fill in the values/words for each `xx` with your answers from question 7 in the one proportion test to create a null distribution with 1000 simulations. Then highlight and run lines 1--10.  **Upload a copy of the created null distribution showing your p-value to Gradescope.**

```{r, echo=TRUE, eval = FALSE}
one_proportion_test(probability_success = xx, # Null hypothesis value
          sample_size = xx, # Enter sample size
          number_repetitions = 1000, # Enter number of simulations
          as_extreme_as = xx, # Observed statistic
          direction = "xx", # Specify direction of alternative hypothesis
          report_value = "proportion") # Reporting proportion or number of successes?
```

**8.  Report the p-value.  Interpret the p-value in context of the study.**

\vspace{1in}

9. Write a conclusion in context of the study.

\vspace{1in}

10. What values should be entered for each of the following into the simulation to create the bootstrap distribution of sample proportions to find a 95\% confidence interval?
\vspace{1mm}

* Sample size:

\vspace{.1in}
 
* Number of successes:
    
\vspace{.1in}
* Number of repetitions:
    
\vspace{.1in}
* Confidence level (as a decimal):
    
\vspace{.1in}

We will use the `one_proportion_bootstrap_CI()` function in `R` (in the `catstats` package) to simulate the bootstrap distribution of sample proportions and calculate a confidence interval. Using the provided `R` script file, fill in the values/words for each `xx` with your answers from question 10 in the one proportion bootstrap confidence interval (CI) code to create a bootstrap distribution with 1000 simulations. Then highlight and run lines 13--16.

```{r, echo=TRUE, eval=FALSE}
one_proportion_bootstrap_CI(sample_size = xx, # Sample size
                    number_successes = xx, # Observed number of successes
                    number_repetitions = 1000, # Number of bootstrap samples to use
                    confidence_level = 0.95) # Confidence level as a decimal
```

11.  Report the 95\% confidence interval.

\vspace{0.5in}

**12. Interpret the 95\% confidence interval.**

\vspace{0.8in}

13. Do the results of the confidence interval agree with the results of the hypothesis test?  Explain your answer.

\vspace{0.8in}

### What does *confidence* mean?

In the interpretation of a 95\% confidence interval, we say that we are 95\% confident that the parameter is within the confidence interval.  Why are we able to make that claim?  What does it mean to say "we are 95\% confident"?

14.  Go to this website, [http://www.rossmanchance.com/ISIapplets.html](http://www.rossmanchance.com/ISIapplets.html) and choose 'Simulating Confidence Intervals'.  In the input on the left-hand side of the screen enter 0.25 for $\pi$, 52 for $n$, and 100 for 'Number of intervals'.  Click 'sample'.
\vspace{1mm}

a. In the graph on the bottom right, click on a green dot.  Write down the confidence interval for this sample given on the graph on the left.  Does this confidence interval contain the null value of 0.25?
\vspace{0.5in}


b. Now click on a red dot.  Write down the confidence interval for this sample.  Does this confidence interval contain the null value of 0.25.?
\vspace{0.5in}


c. How many intervals out of 100 contain $\pi$, the null value of 0.25? *Hint*:  This is given to the left of the graph of green and red intervals.
\vspace{0.5in}

15.  Click on 'sample' nine more times.  Write down the 'Running Total' for the proportion of intervals that contain $\pi$.  

\vspace{0.5in}

**16.  Interpret the level of confidence.  *Hint*: What proportion of samples would we expect to give a confidence interval that contains the parameter of interest?**

\vspace{0.8in}

17. Write a paragraph summarizing the results of the study as if writing a press release.  **Upload your group's paragraph to Gradescope.** Be sure to describe:

* Summary statistic and interpretation

* P-value and interpretation

* Confidence interval and interpretation

* Conclusion (written to answer the research question)

* Scope of inference


\newpage
