## Activity 8b:  Male Boxers --- Theory CI

\setstretch{1}

### Learning objectives

* Calculate a theory-based confidence interval for a single proportion.

* Check the appropriate conditions to find a theory-based confidence interval.

* Interpret a confidence interval for a single proportion.

* Use the normal distribution to find the multiplier needed for a confidence interval

### Terminology review

In this activity, we will introduce simulation-based confidence intervals for a single proportion. Some terms covered in this activity are:

* Parameter of interest

* Multiplier

* Normal distribution

To review these concepts, see Chapter 5 in your textbook, focusing on Sections 5.1 through 5.3.

### Male Boxers

Last class we found very strong evidence that the true proportion of male boxers that are left-handed is greater than 0.1.  In this activity we will use the same data set to find the theory-based 95\% confidence interval.

Remember from the last activity: Left-handedness is a trait that is found in about 10\% of the general population. Past studies have shown that left-handed men are over-represented among professional boxers. The fighting claim states that left-handed men have an advantage in competition.  In this random sample of 500 male professional boxers, we want to see if there is an over-prevalence of left-handed fighters.  In the sample of 500 male boxers, 81 were left-handed.

Recall that to use theory-based methods we must check the conditions to approximate the sampling distribution with the normal distribution.  From the previous activity, we saw that independence was satisfied as the researchers took a random sample.

To check the success-failure condition to use theory-based methods for confidence intervals, we use $\hat{p}$ in the calculations since we are not assuming a value for $\pi$. That is, check that we have at least 10 successes and 10 failures in our **sample**:  $n\hat{p} \geq 0$  and $n(1-\hat{p}) \geq 10$.

1.  Verify that the success-failure condition is met to use theory based methods to find a 95\% confidence interval.  

\vspace{0.5in}
\newpage

To calculate a theory-based 95\% confidence interval for $\pi$, we will first find the **standard error** of $\hat{p}$ by plugging in the value of $\hat{p}$ for $\pi$ in $SD(\hat{p})$:

$$SE(\hat{p}) = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}.$$
Note that we do not include a "0" subscript, since we are not assuming a null hypothesis. 

2.  Calculate the standard error of the sample proportion to find a 95\% confidence interval.

\vspace{0.5in}

Recall from earlier in the semester we learned that the sample standard deviation measures the average variability of the data values in the sample from the sample mean.  In other words, the sample standard deviation measures how far each data point is from the mean of the data, on average.  

3.  Interpret the standard error of the sample proportion found in question 2 in context of the problem.

\vspace{0.8in}


To find the confidence interval, we will add and subtract the **margin of error** to the point estimate:

$$\text{point estimate}\pm\text{margin of error}$$
$$\hat{p}\pm z^* SE(\hat{p})$$

The $z^*$ multiplier is the percentile of a standard normal distribution that corresponds to our confidence level. If our confidence level is 95\%, we find the Z values that encompass the middle 95\% of the standard normal distribution.  If 95\% of the standard normal distribution should be in the middle, that leaves 5\% in the tails, or 2.5\% in each tail.  The `qnorm()` function in `R` will tell us the $z^*$ value for the desired percentile (in this case, 95\% + 2.5\% = 97.5\% percentile). 

```{r, echo=TRUE, collapse=FALSE}
qnorm(0.975) # Multiplier for 95% confidence interval
```

```{r simpleNormaldist, fig.cap="A standard normal curve.", warning=FALSE, out.width="60%"}
X <- seq(-5,5,0.01)
Y <- dnorm(X)
plot(X, Y, type='l', axes=F, xlim=c(-4,4), lwd=2, ylab="", xlab="Z")
#axis(1, at=-3:3)
abline(h=-0.002)
```

4. Fill in the normal distribution above to show how `R` found the $z^*$ multiplier.

\vspace{0.1in}

5.  What is the value of the multiplier needed to calculate the 95\% confidence interval for the true proportion of male boxers that are left-handed?

\vspace{1in}

6.  Calculate the margin of error for the 95\% confidence interval.

\vspace{1in}

7.  Calculate the 95\% confidence interval for the parameter of interest.

\vspace{0.5in}

8.  Interpret the 95\% confidence interval in the context of the problem.

\vspace{1in}

9. Is the null value, 0.1, contained in the 95\% confidence interval?  Explain, based on the p-value from the last activity, why you expected this to be true.
\vspace{1in}

### Simulation Methods {-}

In question 1, we found that the success-failure condition was met to use theory-based methods.  Here we will use simulation methods to find a 95\% confidence interval for the parameter of interest.

Use the `one_proportion_bootstrap_CI()` function in `R` to simulate the bootstrap distribution of sample proportions and calculate a confidence interval. Using the provided `R` script file, fill in the values/words for each `xx` in the one proportion bootstrap confidence interval (CI) code to create a bootstrap distribution with 1000 simulations. 

```{r, echo=TRUE, eval=FALSE}
one_proportion_bootstrap_CI(sample_size = xx, # Sample size
                    number_successes = xx, # Observed number of successes
                    number_repetitions = 1000, # Number of bootstrap samples to use
                    confidence_level = 0.95) # Confidence level as a decimal
```

10. Report the simulation 95\% confidence interval.  Is this confidence interval similar to the confidence interval calculated in question 7?

\vspace{0.8in}


### Effect of Sample Size {-}

Suppose in another sample of 230 male boxers we saw that 37 were left-handed.

11.  Calculate the margin of error for a 95\% confidence interval using a multiplier $z^* = 1.96$ for this sample.  Is the margin of error larger or smaller than the margin of error for the original study?

\vspace{0.5in}

12.  Calculate the 95\% confidence interval for this new study using the margin of error from question 11.  

\vspace{0.5in}

13.  Is the confidence interval calculated in question 11 with the smaller sample size wider or smaller than the confidence interval in question 6?  Explain your answer
\vspace{1in}


### Take-home messages

1.  In theory-based methods, we add and subtract a margin of error to the sample statistic.  The margin of error is calculated using a multiplier that corresponds to the level of confidence times the variability (standard error) of the statistic.

2. The confidence interval calculated using theory-based methods should be similar to the confidence interval found using simulation methods provided the success-failure condition is met.

3.  A smaller sample size will increase the margin of error which results in a wider confidence interval. 

### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity and material covered.


\newpage