## Activity:  IMDb Movie Reviews - Displaying two variables

\setstretch{1}

### Learning objectives

* Identify and create appropriate summary statistics and plots
  given a data set or research question for quantitative data.

* Interpret the following summary statistics in context:
  median, lower quartile, upper quartile,
  standard deviation, interquartile range.

* Given a plot or set of plots, describe and compare the distribution(s)
  of a single quantitative variable
  (center, variability, shape, outliers).

### Terminology review

In this week's activity, we will review summary measures and plots for quantitative variables.  Some terms covered in this activity are:

* Two measures of center: mean, median

* Two measures of spread (variability): standard deviation, interquartile range (IQR)

* Types of graphs: box plots, dot plots, histograms

To review these concepts, see Section 2.3 in the textbook.

### Movies released in 2016

A data set was collected on movies released in 2016.  Here is a list of some of the variables collected on these movies.

| **Variable** 	| **Description** |
|----	|-------------	|
| `budget_mil` | Amount of money (in US $ millions) budgeted for the production of the movie |
| `revenue_mil` | Amount of money (in US $ millions) the movie made after release|
| `duration` | Length of the movie (in minutes)|
| `content_rating` | Rating of the movie (`G`, `PG`, `PG-13`, `R`, `Not Rated`)|
| `imdb_score` | IMDb user rating score from 1 to 10 |
| `genres` | Categories the movie falls into (e.g., Action, Drama, etc.) |
| `movie_facebook_likes` | Number of likes a movie receives on Facebook |

\newpage


#### Displaying a single categorical and single quantitative variable {-}

The box plot of movie budgets (in millions) by content rating is plotted using the code below.  Enter the variable `budget_mil` for yy and the variable `content_rating` for xx at line 31, highlight and run code lines 29 - 35. This plot helps to compare the budget for different levels of content rating.

**Add summary stats**

```{r, out.width="60%", echo=TRUE, eval=FALSE}
movies %>%  # Data set piped into...
  filter(content_rating != "Not Rated") %>% # Remove Not Rated movies
  ggplot(aes(y = yy, x = xx))+  # Identify variables
  geom_boxplot()+  # Tell it to make a box plot
  labs(title = "Side by side box plot of budget by content rating",  # Title
       x = "Content Rating",    # x-axis label
       y = "Budget (in Millions)")  # y-axis label
```

1. Sketch the box plots created using the `R` code.

\vspace{1.5in}


2. Answer the following questions about the box plots created.

   a. Which content rating has the highest center?
\vspace{0.2in}

   b. Which content rating has the largest spread?
\vspace{0.2in}

   c. Which content rating has the most skewed distribution?
\vspace{0.2in}

   d. Fifty percent of movies in 2016 with a PG-13 content rating fall below what value?  What is the name of this value?
\vspace{0.4in}

   e.  What is the value for the third quartile (Q3) for the PG-13 rating?  Interpret this value in context.
\vspace{.8in}


### Out-of-class activity

For a little more practice using `Rstudio` to create graphs of quantitative variables we will look at some other variables in the `Movies` data set.  Download and open the provided `R` script file, highlight and run the first 8 lines of code.

To use the `favstats()` function in the mosaic package with two variables, we will enter the variables as a formula, response~explanatory.

```{r, echo=TRUE, collapse=FALSE}
movies %>% # Data set piped into...
  summarise(favstats(imdb_score~content_rating)) # Apply favstats function to imdb_score
```

Using the provided `R` script file, we will create side-by-side histograms of IMDb by movie content rating. Enter the variable name, `imdb_score` for yy and `content_rating` for xx at line 44, highlight and run lines 39 - 48.

```{r, out.width="60%", echo=TRUE, eval=FALSE}
movies %>%  # Data set piped into...
  filter(content_rating != "Not Rated") %>% # Remove Not Rated movies
  ggplot(aes(y = yy, x = xx))+  # Identify variables
  geom_boxplot()+  # Tell it to make a box plot
  labs(title = "Side by side box plot of budget by content rating",  # Title
       x = "Content Rating",    # x-axis label
       y = "IMDb Score")  # y-axis label
```

1.  Using the provided `favstats` output and the side-by-side box plots, interpret the value of quartile 1 for the R content rating.

\vspace{.8in}

2. Which content rating has the highest center?

\vspace{0.2in}

3. Which variable is the explanatory variable? Response variable?

\vspace{0.4in}