# Inference for Two Categorical Variables: Confidence Intervals

## Reading Guide: Confidence Intervals for a Difference in Proportions

### Section 5.4.3 (Bootstrap confidence interval for a difference in proportions)  {-}

**Videos**  

* 5.4

\setstretch{1.25}

#### Reminders from previous sections {-}

Sample size of group 1: $n_1$

Sample size of group 2: $n_2$

Sample proportion of group 1:  $\hat{p_1}$

Sample proportion of group 2: $\hat{p_2}$

Population proportion of group 1: $\pi_1$

Population proportion of group 2: $\pi_2$

Parameter: a value summarizing a variable(s) for a population

Statistic: a value summarizing a variable(s) for a sample

Point estimate: other names for the statistic from a sample; our best guess for the parameter of interest.

Sampling distribution: plot of statistics from 1000s of samples of the same size taken from the same population.

Standard deviation of a statistic: the variability of statistics from 1000s of samples.  How far, on average, each statistic is from the true value of the parameter.

Standard error of a statistic: estimated standard deviation of a statistic.

Confidence interval: a process to determine how large an effect is; a range of plausible values for the parameter
\rgi Also called ‘estimation’

Margin of error: the value that is added to and subtracted from the sample statistic to create a confidence interval; half the width of a confidence interval

Bootstrapping: the process of drawing with replacement n times from the original sample

Bootstrapped resample: a random sample of size n from the original sample, selected with replacement

Bootstrapped statistic: the statistic recorded from the bootstrapped resample

Confidence level: how confident we are that the confidence interval will capture the parameter


#### Notes {-}
To create a single bootstrap resample for two categorical variables, how many cards will you need and how will the cards be labeled?
\rgs

What is done with the cards once they are labeled?
\rgs

Interpretations of confidence level must include:
\rgs
\rgs

How do you determine if the results of a hypothesis test agree with a confidence interval?
\rgs
\rgs

How are confidence level and significance level related (for a two-sided test)?
\rgs

#### Example: CPR and blood thinner {-}
1.	What is the research question?
\rgs

2.	What is the sample difference in proportions presented in this example?  What notation would be used to represent this value?
\rgs

3.	What is the parameter (using a difference in proportion) representing in the context of this problem?  What notation would be used to represent this parameter?
\rgs
\rgs

4.	How could we use cards to simulate **1** bootstrap resample?  How many blue cards – to represent what?  How many red cards – to represent what?  What would we do with the cards?  What would you record once you have a simulated sample?
\rgs
\rgs

5.	How can we calculate a 90% confidence interval from the bootstrap distribution for this example?
\rgs

6.	What was the 90% confidence interval? 
\rgs

7.	Interpret the confidence interval in the context of the problem.
\rgs
\rgs

8.	Interpret the confidence level in the context of the problem.
\rgs
\rgs

9.	Does the conclusion of the hypothesis test match the confidence interval?
\rgs

### Section 5.4.4 (Theory-based Methods for a Difference in Proportions)  {-}

\setstretch{1}

In section 5.4.4, read only the sub-section on "Confidence interval for $\pi_1 - \pi_2$". The other sections were covered last week.

**Videos**  

* 5.4

\setstretch{1.25}

#### Reminders from previous sections {-}

Central Limit Theorem: For large sample sizes, the sampling distribution of a sample proportion (or mean) will be approximately Normal (bell-shaped and symmetric)

#### Notes {-}

Conditions for the CLT to apply for two categorical variables
1.	Independence: 
\rgs

\rgi a.	Checked by: 
\rgs

2.	Success/Failure condition: 
\rgs

\rgi a.	Checked by: 
\rgs

#### Formulas {-}

$SD(\hat{p_1} - \hat{p_2})=$
\rgs

Standard error of the difference in sample proportions when we do not assume the null hypothesis is true:
$SE(\hat{p_1} - \hat{p_2})=$
\rgs

Theory-based confidence interval for a s difference in sample proportions: 
\rgs

Margin of error of a confidence interval for a difference in sample proportions: 
\rgs


#### Example: CPR and blood thinner {-}

1. What is the sample difference in proportions presented in this example?  What notation would be used to represent this value?
\rgs

2. What is the parameter (using a difference in proportion) representing in the context of this problem?  What notation would be used to represent this parameter?
\rgs

3. Calculate the standard error of the difference in sample proportions without assuming a null hypothesis.
\rgs

4. Calculate the 90% confidence interval using z^⋆=1.65 as the multiplier.
\rgs

\rgi *Note: a confidence interval interpretation and confidence level interpretation for this example can be found in the Reading Guide solutions for 5.4.1to5.4.2 and 5.4.3.*


\newpage

## Activity:  Winter Sports Helmet Use and Head Injuries - Estimation

\setstretch{1}

### Learning objectives

* Assess the conditions to use the standard normal distribution for a difference in proportions.

* Create and interpret a confidence interval for the difference in proportions.

### Terminology review
In this week's activity, we will use theory-based methods to estimate the difference in two proportions. Some terms covered in this activity are:

* Standard normal distribution

* Independence and success-failure conditions

To review these concepts, see Chapter 5 in your textbook.

### Helmet use and head injuries

In activity 8, we found very strong evidence that helmet use is associated with a reduction in head injury for skiers and snowboarders.  In this in-class activity we will estimate the parameter of interest, the difference in true proportion of skiers and snowboarders with head injuries for those who wore helmets and those that did not, using theory-based methods.


The summary results from a random sample of 3562 skiers and snowboarders involved in accidents is displayed in the two-way table below. Estimate the difference in true proportion of skiers and snowboarders with head injuries for those who wore helmets and those that did not. 

|                | Helmet Use | No Helmet Use | Total |
|:--------------:|:----------:|:-------------:|:-----:|
| Head Injury    |     96     |      480      |  576  |
| No Head Injury |     656    |      2330     |  2986 |
| Total          |     752    |      2810     |  3562 |

#### Vocabulary review. Complete Q1$-$Q3 before class. {-}

1.  Report the point estimate for this study.  Use helmet use minus no helmet use as the order of subtraction.

\vspace{0.4in}

2. Write the conclusion for the study from activity 8.

\vspace{1in}

3. Based on the results from activity 8, do you expect the 95\% confidence interval to contain the null value of zero?  Explain your answer.

\vspace{1in}


#### Use statistical analysis methods to draw inferences from the data {-}

In this activity we will focus on theory-based methods to calculate a confidence interval.  Like with a single proportion, the difference in proportions can be mathematically modeled using the normal distribution if certain conditions are met.

Conditions for the sample distribution of $\hat{p}_1-\hat{p}_2$:

* Independence: The data are independent within and between the two groups.

* Success-Failure Condition: For a confidence interval, the success-failure condition holds for each group.  All cells in the table have at least 10 observations.


4. In activity 8, we saw that the independence condition was met since the study used a random sample.  Is the success-failure condition to find the confidence interval met for each group?  Explain your answer.

\vspace{1in}

To find a confidence interval for the difference in proportions we will add and subtract the margin of error from the point estimate to find the two endpoints.

 $$\hat{p}_1-\hat{p}_2\pm z^*SE(\hat{p}_1-\hat{p}_2), \text{where}$$
 
 $$SE(\hat{p}_1-\hat{p}_2) = \sqrt{\left(\frac{\hat{p}_1 (1-\hat{p}_1)}{n_1}+\frac{\hat{p}_2 (1-\hat{p}_2)}{n_2}\right)}$$
 
Note that the formula changes when calculating the variability around the statistic in order to calculate a confidence interval from the formula used in activity 8!  Here, use the sample proportions for each group to calculate the standard error for the difference in proportions.


5. Calculate the standard error for a difference in proportions to create a 95\% confidence interval.  

\vspace{1in}

6. Interpret the value calculated in question 5 in context of the problem.

\vspace{1in}

The $z^*$ multiplier is found under the standard normal distribution. We find the values that encompass the middle 95% of the distribution.  If 95\% of the standard normal distribution should be in the middle, that leaves 5\% in the tails, or 2.5\% in each tail.  The qnorm function in `R` will tell us the $z^*$ value for the desired percentile (in this case, 95\% + 2.5\% = 97.5\% percentile). 

```{r, echo=TRUE, collapse=FALSE}
qnorm(0.975) # Multiplier for 95% confidence interval
```

7. Sketch a graph to explain how the `R` code above is used to find the $z^*$ multiplier.  

\vspace{1.5in}

8. Using the multiplier of $z^*$ = 1.96 and the standard error found in question 5, calculate the margin of error for a 95\% confidence interval.

\vspace{0.5in}

9. Calculate the 95\% confidence interval for the difference in true proportion of head injuries for those that used helmets minus those who did not. 

\vspace{1in}

10. Interpret the confidence interval found in question 9 in context of the problem.

\vspace{1in}

### Effect of sample size 

Suppose in another sample of skiers and snowboarders involved in accidents we saw these results:

|                | Helmet Use | No Helmet Use | Total |
|:--------------:|:----------:|:-------------:|:-----:|
|   Head Injury  |     135    |      674      |  809  |
| No Head Injury |     921    |      3270     |  4191 |
|      Total     |    1056    |      3944     |  5000 |


11.  Calculate the margin of error for a 95\% confidence interval using a multiplier of $z^*$ = 1.96 for this new sample.  Is the margin of error larger or smaller than the margin of error for the original study?

\vspace{1in}

12.  Calculate the 95\% confidence interval for this new study using the margin of error from question 11.  

\vspace{1in}

13.  Is the confidence interval calculated in question 12 with the larger sample size, wider or smaller than the confidence interval in question 9?

\vspace{1in}

\newpage

### Relative risk

Another summary statistic that can be calculated for two categorical variables is the relative risk.  The relative risk is calculated as the ratio of the conditional proportions.


$$\text{relative risk} = \frac{\hat{p}_1}{\hat{p}_2}$$

14.  Calculate the relative risk of head injury for those who wore helmets compared to those who did not.

\vspace{1in}

15.  Interpret the relative risk in context of the problem.

\vspace{1in}




### Out-of-class activity

The remaining questions cover simulation methods for creating the bootstrap distribution to find the confidence interval. Use section 5.4.3 in the textbook and the TwoPropSim video to complete the following questions.

To create the bootstrap distribution, we will use the two proportion bootstrap CI function.  We will need to enter the response variable name and the explanatory variable name for the formula, the data set name identified above as `injury`, the outcome for the explanatory variable to give the order of subtraction for First in Subtraction, number of repetitions, the outcome for the response variable that is a success for Response Value Numerator, and the confidence level as a decimal.

The response variable name is Outcome and the explanatory variable name is Helmet.

\newpage
1. What values should be entered for each of the following into the simulation to create a 95\% confidence interval?

\vspace{.5mm}
* First in Subtraction (What is the outcome for the explanatory variable that is used as first in the order of subtraction? `"Yes"` or `"No"`):

\vspace{.2in}
* Response Value Numerator (What is the outcome for the response variable that is considered a success? `"Head Injury"` or `"No Head Injury"`):

\vspace{.2in}
* Number of repetitions:

\vspace{.2in}
* Confidence Level (entered as a decimal):

\vspace{.2in}

Using the `R` script file for this activity, enter your answers for question 1 in place of the xx's in the two proportion bootstrap CI function to produce the bootstrap distribution with 1000 simulations, highlight and run lines 1$-$22.

```{r, echo=TRUE, eval=FALSE}
two_proportion_bootstrap_CI(formula = Outcome ~ Helmet, 
                            data=injury, # Name of dataset
                            first_in_subtraction = "xx", # Order of subtraction: enter the name of Group 1
                            response_value_numerator = "xx", # Define which outcome is a success 
                            number_repetitions = 1000, # Always use a minimum of 1000 repetitions
                            confidence_level = 0.95) # Enter the level of confidence as a decimal

```

2. Report the bootstrap 95\% confidence interval. 

\vspace{0.5in}

3. What percentile does the upper value of the confidence interval represent?

\vspace{0.3in}

4. Is the bootstrap 95\% confidence interval similar to what was found using theory-based methods?  Why did you expect this to be true?

\vspace{0.5in}

### Take-home messages

1.	Simulation methods and theory-based methods should give the same results for a study if the validity conditions are met.  Check the validity conditions for each type of test to determine if theory-based methods can be used.

2.  When calculating the standard error for the difference in proportions when doing a hypothesis test we use the pooled proportion of successes, the best estimate for calculating the variability under the assumption the null hypothesis is true.  For a confidence interval, we use the values of the two conditional proportions to calculate the standard error.  Make note of the difference in these two formulas. 

3.  In addition to estimating the difference in proportions for two categorical variables we can also find the relative risk, the ratio of conditional proportions.  

4.  Increasing sample size will result in less sample to sample variability which will result in a narrower confidence interval.  


### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity.
