## Week 10 Lab:  Diabetes

\setstretch{1}

### Learning objectives

* Given a research question involving two categorical variables, construct the null and alternative hypotheses
  in words and using appropriate statistical symbols.
  
* Assess the conditions to use the normal distribution model for a difference in proportions.

* Calculate the Z test statistic for a difference in proportions.

* Find, interpret, and evaluate the p-value for a theory-based hypothesis test for a difference in proportions.

* Create and interpret a theory-based confidence interval for a difference in proportions.

### Glycemic Control in Diabetic Adolescents

Researchers compared the efficacy of two treatment regimens to achieve durable glycemic control in children and adolescents with recent-onset type 2 diabetes.  A convenience sample of patients 10 to 17 years of age with recent-onset type 2 diabetes were randomly assigned to either a medication (rosiglitazone) or a lifestyle-intervention program focusing on weight loss through eating and activity.  Researchers measured whether the patient still needs insulin (failure) or had glycemic control (success).  Of the 233 children who received the Rosiglitazone treatment,143 had glycemic control, while of the 234 who went through the lifestyle-intervention program, 125 had glycemic control. Is there evidence that there is difference in proportion of patients that achieve durable glycemic control between the two treatments?  Use Rosiglitazone – Lifestyle as the order of subtraction.

Upload and open the `R` script file for Week 10 lab. Upload and import the csv file, `diabetes`. Enter the name of the data set (see the environment tab) for `datasetname` in the R script file in line 5. Highlight and run lines 1--6 to get the counts for each combination of categories.

```{r, out.width="60%", echo=TRUE, eval=FALSE}
datasetname -> diabetes
diabetes %>% group_by(treatment) %>% count(outcome)
```


1. Is this an observational study or a randomized experiment?  Explain your answer. 
\vspace{0.5in}

2. Complete the following two-way table using the `R` output.

|                         |     Rosiglitazone    |     Lifestyle    |     Total    |
|-------------------------|----------------------|------------------|--------------|
|     Glycemic Control    |                      |                  |              |
|     Insulin Required    |                      |                  |              |
|     Total               |                      |                  |              |



3.   **Write out the parameter of interest for this study.**
\vspace{0.5in}

4.  Using the research question, write the alternative hypothesis in notation.
\vspace{1in}

5. Is the independence condition met for this study? Explain your answer.
\vspace{0.5in}

6.  Calculate the summary statistic.  Use proper notation.
\vspace{0.5in}

Fill in the missing values/names in the `R` script file in the two-proportion_test function to create the null distribution and find the simulation p-value for the test.

```{r, echo=TRUE, eval=FALSE}
two_proportion_test(formula = outcome~treatment, # response ~ explanatory
    data= diabetes, # Name of data set
    first_in_subtraction = "xx", # Order of subtraction: enter the name of Group 1
    number_repetitions = 1000, # Always use a minimum of 1000 repetitions
    response_value_numerator = "xx", # Define which outcome is a success 
    as_extreme_as = xx, # Calculated observed statistic (difference in sample proportions)
    direction="xx") # Alternative hypothesis direction ("greater","less","two-sided")
```

7. Report the simulation p-value.
\vspace{0.2in}

8.  **Would the theory-based p-value be similar to the simulation p-value?  Explain your answer.**
\vspace{0.5in}

9. **Calculate the number of standard errors the sample difference in proportions is from the null value of zero?**
\vspace{0.8in}

10.  **Would a simulation 95\% confidence interval contain the null value of zero?  Explain your answer.**
\vspace{0.8in}

11. Calculate the standard error for a difference in proportions to create a 95\% confidence interval.  

\vspace{1in}

```{r, echo=TRUE, collapse=FALSE}
qnorm(0.975) # Multiplier for 95% confidence interval
```

12. **Using the multiplier of $z^*$ = 1.96 and the standard error found in question 11, calculate the 95\% confidence interval for the parameter of interest.** 

\vspace{1in}

13. Interpret the confidence interval found in question 12 in context of the problem.

\vspace{1in}

14.  Write a conclusion to the research question.
\vspace{0.8in}

15. Write a paragraph summarizing the results of the study.  **Upload a copy of your group's summary paragraph to Gradescope.** Be sure to describe:

* Summary statistic and interpretation

* P-value and interpretation

* Confidence interval and interpretation

* Conclusion (written to answer the research question)

* Scope of inference

\vspace{2in}

