## Activity 12b:  Construction Costs

\setstretch{1}

### Learning outcomes

* Given a research question involving paired differences, construct the null and alternative hypotheses
  in words and using appropriate statistical symbols.
  
* Describe and perform a theory-based hypothesis test for a paired mean difference.

* Interpret and evaluate a p-value for a theory-based hypothesis test for a paired mean difference.

* Use theory-based methods to find a confidence interval for a paired mean difference.

* Interpret a confidence interval for a paired mean difference.

* Use a confidence interval to determine the conclusion of a hypothesis test.

### Terminology review

In this week's activity, we will analyze paired quantitative data using simulation-based methods. Some terms covered in this activity are:

* Paired data

* Mean difference

* Independent observational units

* Normality

* $t$-distribution

* Degrees of freedom

* T-score

To review these concepts, see Sections 6.1 and 6.2 in the textbook.

### Constructions costs

The year 2021 has seen massive fluctuations in construction costs.  Lumber futures alone soared from roughly 700 USD/1,000 board feet in January to a peak of 1711 USD/1,000 board feet in May, only to plummet back to 500 USD/1,000 board feet by August. Researchers would like to determine how these wildly variable costs have impacted the accuracy of construction cost estimates.  They collect a random sample of 372 construction projects from 20 different U.S. zip codes to investigate if current construction cost estimates (in 1,000 U.S. Dollars) match, on average, the actual construction cost (in 1,000 U.S. dollars). 

#### Ask a research question {-}

1. Write out the null hypothesis in notation for this study.  Be sure to clearly identify the subscripts.

\vspace{0.5in}

2. Write out the alternative hypothesis in words for this study.

\vspace{0.8in}

The sampling distribution for $\bar{x}$ based on a sample of size $n$ from a population with a true mean $\mu$ and true standard deviation $\sigma$ can be modeled using a normal distribution when certain conditions are met.

Conditions for the sampling distribution of $\bar{x}$ to follow an approximate normal distribution:

* **Independence**: The sample’s observations are independent

* **Normality**: The data should be approximately normal or the sample size should be large.

    - $n < 30$: If the sample size $n$ is less than 30 and there are no clear outliers in the data, then we typically assume the data come from a nearly normal distribution to satisfy the condition.

    - $n \geq 30$: If the sample size $n$ is at least 30 and there are no particularly extreme outliers, then we typically assume the sampling distribution of $\bar{x}$ is nearly normal, even if the underlying distribution of individual observations is not.
    
    ```{r, echo=TRUE}
SUD <- read.csv("https://math.montana.edu/courses/s216/data/SUD.csv")
SUD %>%  # Data set piped into...
  ggplot(aes(y = stay, x = group))+  # Identify variables
  geom_boxplot()+  # Tell it to make a box plot
  labs(title = "Length of Stay with and without SUD Treatment",  # Title
       x = "Group",    # x-axis label
       y = "Length of Stay (days)")  # y-axis label
```
    

The following code gives the summary statistics for the data.

```{r, echo=TRUE}
SUD %>%
  summarize(favstats(stay ~ group))
```

1.  Verify that the independence condition was satisfied.  
\vspace{0.8in}

2. Is the normality condition met to use the theory-based methods for analysis?  Explain your answer.
\vspace{1in}


#### Summarize and visualize the data {-}

5. Calculate the summary statistic for this study.  Use treated minus untreated as the order of subtraction.  Give the appropriate notation with cleary defined subscripts.

\vspace{0.8in}

#### Use statistical inferential methods to draw inferences from the data {-}


To find the standardized statistic for the paired differences we will use the following formula:

$$T = \frac{\bar{x}_d}{SE(\bar{x}_d)},$$

where the standard error of the sample mean difference is:

$$SE(\bar{x}_d)=\frac{s_d}{\sqrt{n}}.$$

2.  Calculate the standard error of the sample mean difference.

\vspace{0.5in}

3.  Calculate the standardized statistic.

\vspace{0.5in}

Using the provided `R` script file, enter the T-score (for `xx`) into the `pt()` function using a `df` = $n_d-1 = 33 - 1 = 32$, and `lower.tail = TRUE` to find the p-value.  Highlight and run line 31.  

```{r, echo=TRUE, eval=FALSE}
pt(xx, df=32, lower.tail=TRUE)
```
10. Explain why we multiplied by 2 in the code above.
\vspace{0.3in}

11.  Interpret the p-value in context of the study.
\vspace{0.8in}


12.  Do you expect the 95\% confidence interval to contain the null value of zero?  Explain your answer.
\vspace{0.8in}

4.  Is the p-value found using theory-based methods similar to the simulation p-value found in the in-class activity?

\vspace{0.5in}


To calculate the 99\% theory-based confidence interval for the paired mean difference, use the following formula:

$$\bar{x}_d\pm t^* SE(\bar{x}_d).$$

We will need to find the $t^*$ multiplier using the function `qt()`. For a 99\% confidence level, we are finding the $t^*$ value at the 99.5th percentile with `df` = $n_d - 1 = 33 - 1 = 32$.

```{r, echo=TRUE}
qt(0.995, df = 32, lower.tail=TRUE)
```

5.  Calculate the 99\% confidence interval for the paired mean difference using theory-based methods.

\vspace{1in}


15.  Interpret the 95\% confidence interval in context of the study.

\vspace{1in}

16.  Do the results of the CI agree with the p-value?  Explain your answer.

\vspace{0.5in}

17.  Write a conclusion to the test in context of the study.
\vspace{0.8in}

### Take-home messages

1.  In order to use theory-based methods for dependent groups (paired data), the independent observational units and normality conditions must be met.  

2.  A T-score is compared to a $t$-distribution with $n - 1$ df in order to calculate a one-sided p-value. To find a two-sided p-value using theory-based methods we need to multiply the one-sided p-value by 2.  

3.  A $t^*$ multiplier is found by obtaining the bounds of the middle X% (X being the desired confidence level) of a $t$-distribution with $n - 1$ df.


### Additional notes

Use this space to summarize your thoughts and take additional notes on today's activity and material covered

\newpage
