## Week 12 Lab:  Swearing

\setstretch{1}


### Learning objectives

* Identify whether a study is a paired design or independent groups

* Given a research question involving paired data, construct the null and alternative hypotheses
  in words and using appropriate statistical symbols.

* Describe and perform a simulation-based hypothesis test for a mean difference.

* Interpret and evaluate a p-value for a hypothesis test for a mean difference.

* Use bootstrapping methods to find a confidence interval for a mean difference.

* Interpret a confidence interval for a mean difference.

### Type of samples

For each of the following scenarios, determine whether the samples are paired or independent.

1.

2. **GS **

### Swearing

Profanity (language considered obscene or taboo) and society's attitude about its acceptableness is a highly debated topic, but does swearing serve a physiological purpose or function? Previous research has shown that swearing produces increased heart rates and higher levels of skin conductivity. It is theorized that since swearing provokes intense emotional responses, it acts as a distracter, allowing a person to withstand higher levels of pain. To explore the relationship between swearing and increased pain tolerance, researchers from Keele University (Staffordshire, UK) recruited 83 native English-speaking participants. Each volunteer performed two trials holding a hand in an ice-water bath, once while repeating the “f-word” every three seconds, and once while repeating a neutral word (“table”). The order of the word to repeat was randomly assigned. Researchers recorded the length of time, in seconds, from the moment the participant indicated they were in pain until they removed their hand from the ice water for each trial. They hope to find evidence that pain tolerance is greater (longer times) when a person swears compared to when they say a neutral word, on average. Use Swear – Neutral as the order of subtraction.

4. What does $\mu_d$ represent in the context of this study?

\vspace{0.8in}

5.  Write out the null hypothesis in words for this study.  

\vspace{0.8in}

5. Write out the alternative hypothesis in proper notation for this study.

\vspace{0.5in}

Upload and open the `R` script file for Week 12 lab. Upload and import the csv file, `swearing`. Enter the name of the data set (see the environment tab) for datasetname in the `R` script file in line 7. Write a title for the boxplots in line 11. Highlight and run lines 1--13 to load the data and create plots of the data.

```{r, echo=TRUE, eval=FALSE}
datasetname <- swearing 
swearing %>%  # Data set piped into...
  ggplot(aes(y = Baseweight, x = Trail))+  # Identify variables
  geom_boxplot()+  # Tell it to make a box plot
  labs(title = "xx",  # Title
       x = "Trail",    # x-axis label
       y = "Baseweight(lbs)")  # y-axis label
```

6. Based on the plots, which trail has the highest median baseweight?
\vspace{0.2in}

Enter the name of the explanatory variable for `explanatory` and the name of the response variable for `response` in line 17.  Highlight and run lines 16--17 to get the summary statistics for the data.

```{r, echo=TRUE, eval=FALSE}
hikes %>%
  summarize(favstats(response~explanatory))
```
7. **Calculate the summary statistic for this study.  Use appropriate notation.**

\vspace{1in}

### Use statistical inferential methods to draw inferences from the data {-}

8.  **Using the provided graphs and summary statistics, determine if it would make the most sense to use theory-based methods or simulation methods.  Explain your reasoning.**

\vspace{0.8in}

### Hypothesis test {-}

Remember that the null distribution is created based on the assumption the null hypothesis is true.  In this study, the null hypothesis states that there is no association between the two variables.  This means that the base weight observed in the data set would have been the same regardless of the type of trails.

We will use the `two_mean_test()` function in `R` (in the `catstats` package) to simulate the null distribution of differences in sample means and compute a p-value. 

9.  When using the `two_mean_test()` function, we need to enter the name of the response variable, `Baseweight`, and the name of the explanatory variable, `Trail`, for the formula.  The name of the data set as shown above is `hikes`.  What values should be entered for each of the following to create 1000 simulated samples?

\vspace{1mm}
* First in subtraction (What is the outcome for the explanatory variable that is used as first in the order of subtraction? `"CDT"` or `"PCT"`):

\vspace{.2in}
* Number of repetitions:
    
\vspace{.2in}
* As extreme as:
    
\vspace{.2in}
* Direction (`"greater"`, `"less"`, or `"two-sided"`):

\vspace{.2in}

10.  Simulate a null distribution and compute the p-value. Using the `R` script file for this activity, enter your answers for question 9 in place of the `xx`'s to produce the null distribution with 1000 simulations.  Highlight and run lines 20--25.  **Upload a copy of the null distribution created to Gradescope for your group.**

```{r, echo = TRUE, eval = FALSE}
two_mean_test(Baseweight~Trail, 
         data = hikes,  # Variables and data
         first_in_subtraction = "xx", # First outcome in order of subtraction
         number_repetitions = 1000,  # Number of simulations
         as_extreme_as = xx,  # Observed statistic
         direction = "xx")  # Direction of alternative: "greater", "less", or "two-sided"
```

|        Sketch the null distribution created using the code above.
\vspace{1.5in}

### Communicate the results and answer the research question {-}

11.  **Report the p-value. Based off of this p-value, write a conclusion to the hypothesis test.**

\vspace{0.8in}

12. Do you expect the 95\% confidence interval to contain the null value of zero?  Explain.

\vspace{0.8in}

### Confidence interval {-}

We will use the `two_mean_bootstrap_CI()` function in `R` (in the `catstats` package) to simulate the bootstrap distribution of differences in sample means and calculate a confidence interval. 

13. Using bootstrapping find a 95\% confidence interval. Using the provided `R` script file, enter the variable names and data set name as in the `two_mean_test()` function, outcome name for the first in subtraction, number of repetitions, and the confidence level as a decimal.  Highlight and run lines 28--32. Report the 95\% confidence interval in interval notation.
```{r, echo = TRUE, eval = FALSE}
two_mean_bootstrap_CI(RESPONSE ~ EXPLANATORY, 
                      data = DATASET,  # Variables and data
                      first_in_subtraction = "xx", # First value in order of subtraction
                      number_repetitions = 1000,  # Number of simulations
                      confidence_level = xx)
```

\vspace{0.3in}

14. Interpret the interval you calculated in question 13. 

\vspace{0.8in}

24.  Write a paragraph summarizes the results of this study as if you were describing the results to your roommate.  Be sure to describe:

* Summary statistic

* P-value and interpretation

* Conclusion (written to answer the research question)

* Confidence interval and interpretation

* Scope of inference

\vspace{2.6in}

\newpage